---
title: "Kitchener Landmarks Dashboard"
author: "Hiral Bhavsar"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    theme: cosmo
---


```{r setup, include=FALSE}
# Load required libraries
if (!requireNamespace("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load("flexdashboard", "leaflet", "sf", "dplyr", "ggplot2", "plotly", 
               "DT", "shiny", "tidyr", "ggrepel")

# Load Landmark Data:
arcgis_url <- "https://services1.arcgis.com/qAo1OsXi67t7XgmS/arcgis/rest/services/Landmarks/FeatureServer/0/query?outFields=*&where=1%3D1&f=geojson"
landmark_data <- st_read(arcgis_url, quiet = TRUE)

# Convert timestamps to standard date format:
landmark_data <- landmark_data |>
  mutate(
    CREATE_DATE = as.Date(as.POSIXct(CREATE_DATE / 1000, origin="1970-01-01")),
    UPDATE_DATE = as.Date(as.POSIXct(UPDATE_DATE / 1000, origin="1970-01-01")),
    SOURCE_DATE = as.Date(as.POSIXct(SOURCE_DATE / 1000, origin="1970-01-01")),
    STATUS = as.factor(STATUS),
    CATEGORY = as.factor(CATEGORY)
  )

# To verify numeric and factor conversion:
landmark_data <- landmark_data |>
  mutate(
    X_COORD = as.numeric(X_COORD),
    Y_COORD = as.numeric(Y_COORD),
    WARDID = as.numeric(WARDID),
    PLANNINGCOMMUNITYID = as.numeric(PLANNINGCOMMUNITYID),
    SUBCATEGORY = as.factor(SUBCATEGORY),
    MUNICIPALITY = as.factor(MUNICIPALITY),
    DIVISION_RESPONSIBLE = as.factor(DIVISION_RESPONSIBLE),
    OWNERSHIP = as.factor(OWNERSHIP)
  )

# To handle missing values:
landmark_data <- landmark_data |>
  mutate(
    CIVIC_NO = ifelse(is.na(CIVIC_NO), 0, CIVIC_NO),
    UNIT_NO = ifelse(is.na(UNIT_NO), "Unknown", UNIT_NO),
    BUILDINGID = ifelse(is.na(BUILDINGID), 0, BUILDINGID)
  )

# Remove columns which has missing or no data:
landmark_data <- subset(landmark_data, select = -c(OBJECTID, HYPERLINK))

# To ensure geometry is POINT type:
landmark_data <- st_cast(landmark_data, "POINT") |>
  mutate(Longitude = st_coordinates(geometry)[,1], Latitude = st_coordinates(geometry)[,2])

# Classification of landmarks into five categories: 
landmark_data$Class <- case_when(
  landmark_data$CATEGORY %in% c("PARK", "PLAYGROUND", "DOG PARK", "TOBOGGAN HILL", "GOLF COURSE") ~ "Recreational Spaces",
  landmark_data$CATEGORY %in% c("SPORTS SOCCER", "SPORTS BASKETBALL", "SPORTS TENNIS COURT", "SPORTS CRICKET", "SPORTS BALLDIAMOND", "SPORTS ARENA", "SPORTS POOL", "SPORTS MISC") ~ "Sports Facilities",
  landmark_data$CATEGORY %in% c("LIBRARY", "LITTLE LIBRARY", "GOVERNMENT SERVICE", "EMERGENCY SERVICE", "SENIOR CARE FACILITY", "VULNERABLE SERVICES") ~ "Community & Public Services",
  landmark_data$CATEGORY %in% c("PLACE OF WORSHIP", "ART AND ARTIFACTS", "INDIGENOUS RESOURCES", "CEMETERY") ~ "Cultural & Religious Sites",
  landmark_data$CATEGORY %in% c("BUSINESS PARK", "OPERATIONS CENTRE", "COMMUNITY RESOURCE", "EDUCATION FACILITY") ~ "Urban Infrastructure",
  TRUE ~ "Other"
)
```

## ðŸ“‹ Landmark Overview {.tabset}

### ðŸ¢ Landmark Data Table

```{r}
datatable(landmark_data |>
            select(LANDMARK, Class, CATEGORY, MUNICIPALITY, OWNERSHIP, WARDID, CREATE_DATE),
          options = list(pageLength = 10, scrollX = TRUE, autoWidth = TRUE, searchHighlight = TRUE),
          rownames = FALSE)
```

### ðŸ“Š Landmark Distribution

```{r}
# Define specific colors to individual class:
real_world_colors <- c(
  "Community & Public Services" = "steelblue",  # Blue for government/public services
  "Cultural & Religious Sites" = "forestgreen",  # Green for cultural/religious places
  "Other" = "gray",  # Gray for unspecified categories
  "Recreational Spaces" = "purple",  # Purple for parks/playgrounds
  "Sports Facilities" = "red",  # Red for sports
  "Urban Infrastructure" = "darkorange" # Orange for city infrastructure
)

landmark_counts <- landmark_data |>
  group_by(Class, CATEGORY) |>
  summarise(Total = n(), .groups = "drop")

landmark_plot <- ggplot(landmark_counts, aes(
  x = reorder(Class, -Total),  
  y = Total,
  fill = Class,
  text = paste("Class:", Class, "<br>Category:", CATEGORY, "<br>Total:", Total)
)) +
  geom_bar(stat = "identity") +  
  coord_flip() +
  scale_fill_manual(values = real_world_colors) +  # âœ… Apply real-world colors here
  theme_minimal() +
  labs(
    title = "Interactive Landmark Distribution by Class",
    x = "Class",
    y = "Number of Landmarks",
    fill = "Landmark Class",
    caption = "This chart represents the total count of landmarks categorized by Class and Category."
  ) +
  theme(legend.position = "right")

ggplotly(landmark_plot, tooltip = "text")

```

### ðŸ“‰ Time Series Analysis
```{r}
# Extract Year and count Landmarks by Class:
landmark_trends <- landmark_data |>
  mutate(Year = lubridate::year(CREATE_DATE)) |>
  filter(!is.na(Year)) %>%
  count(Year, Class) %>%
  arrange(Year)


# Line Chart:
trend_chart <- ggplot(landmark_trends, aes(x = Year, y = n, color = Class)) +
  geom_point(size = 2, alpha = 0.8) +  # Show actual data points
  geom_line(size = 1, alpha = 0.8) +  # Connect the actual data points
  labs(title = "Landmark Creation Trends Over Time",
       subtitle = "3-Year Moving Average Applied for Clarity",
       x = "Year", 
       y = "Total Landmarks",
       color = "Landmark Class") +
  scale_x_continuous(breaks = seq(min(landmark_trends$Year), max(landmark_trends$Year), by = 3)) +  # Reduce x-axis clutter
  scale_y_continuous(labels = scales::comma) +  # Format y-axis numbers
  scale_color_manual(values = real_world_colors) +  # Apply real-world colors
  theme_minimal() +
  theme(legend.position = "right")

# Convert to Interactive Plot:
ggplotly(trend_chart)

```
### ðŸ› Ownership Distribution 

```{r}
# Define ownership categories:
selected_ownership <- c("KITCHENER", "PRIVATE")

# Ensure OWNERSHIP is consistently formatted:
filtered_data <- landmark_data |>
  mutate(OWNERSHIP = toupper(OWNERSHIP)) |>  # Convert to uppercase
  filter(OWNERSHIP %in% selected_ownership)

# Count total landmarks by ownership type:
ownership_count <- filtered_data |>
  count(OWNERSHIP)

# Calculate Ratio:
kitchener_count <- ownership_count$n[ownership_count$OWNERSHIP == "KITCHENER"]
private_count <- ownership_count$n[ownership_count$OWNERSHIP == "PRIVATE"]

if (!is.na(kitchener_count) & !is.na(private_count) & private_count > 0) {
  ownership_ratio <- round(kitchener_count / private_count, 1)  # Calculate how many times more
} else {
  ownership_ratio <- "N/A"  # Handle division by zero case
}

# Modify Ownership Labels to Include Ratio (for the legend only):
ownership_count <- ownership_count |>
  mutate(OWNERSHIP_DISPLAY = case_when(
    OWNERSHIP == "KITCHENER" ~ paste0("KITCHENER (", ownership_ratio, "x more than PRIVATE)"),
    TRUE ~ OWNERSHIP
  ))

# Create Bar Chart with Custom Tooltip:
ownership_bar_chart <- ggplot(ownership_count, aes(x = OWNERSHIP_DISPLAY, y = n, fill = OWNERSHIP_DISPLAY, 
                                                   text = paste0("Ownership: ", OWNERSHIP, 
                                                                 "<br>Number of Landmarks: ", n))) +
  geom_bar(stat = "identity", width = 0.6) +
  labs(title = "Total Landmarks Owned by Kitchener vs Private",
       x = "Ownership Type", 
       y = "Number of Landmarks",
       fill = "Ownership") +
  theme_minimal() +
  theme(legend.position = "top")  # Move legend to top

# Convert to Interactive Plot with Custom Tooltip:
ggplotly(ownership_bar_chart, tooltip = "text")


```

### Spatial Distribution (Map)
```{r}
# Define color palette for Class distribution
color_palette <- colorFactor(
  palette = real_world_colors,
  domain = landmark_data$Class
)

# Create the Leaflet map
leaflet(landmark_data) |>
  addTiles() |>
  addCircleMarkers(
    lng = ~Longitude, 
    lat = ~Latitude,
    radius = 3,
    color = ~color_palette(Class),
    popup = ~paste0(
      "<strong>Landmark: </strong>", LANDMARK, "<br>",
      "<strong>Class: </strong>", Class, "<br>",
      "<strong>Address: </strong>", STREET, "<br>",
      "<strong>Created on: </strong>", CREATE_DATE, "<br>",
      # To add image in the popup:
     "<br/><img src = 'https://encrypted-tbn0.gstatic.com/images?q=tbn:ANd9GcQUKAYuOjghdYK7J3p_Z27HOBjg04p0lAILPA&s' width = '150px'>",
     # To add location link in the popup:
      "<br/><a href='https://www.openstreetmap.org/?mlat=", Latitude, "&mlon=", Longitude, "' target='_blank'>View on OpenStreetMap</a><br>",
     # To add source(reference) link in the popup:
      "<a href='https://open-kitchenergis.opendata.arcgis.com/datasets/5dbeed2b0f5e4905934436709b4ac7f8_0/explore' target='_blank'>Source of Data</a>"
    )
  ) |>
  addLegend(
    "bottomright",
    pal = color_palette,
    values = ~Class,
    title = "Landmark Class Distribution",
    opacity = 1
  )



